---
title: "Animations in ggplot - gganimate"
author: "Gareth Green"
output: slidy_presentation
---

```{r echo = FALSE}
# Course: 5210 Communicating Data
# Purpose: Demonstrate gganimate
# Date: August 5, 2019
# Author: Gareth Green

```

```{r echo = FALSE}
# Clear environment of variables and functions
rm(list = ls(all = TRUE)) 

# Clear environmet of packages
if(is.null(sessionInfo()$otherPkgs) == FALSE)lapply(paste("package:", names(sessionInfo()$otherPkgs), sep=""), detach, character.only = TRUE, unload = TRUE)

```

```{r echo = FALSE, message = FALSE, warning = FALSE}
# Load libraries
library(tidyverse)
library(gganimate)

```

```{r echo = FALSE}
# Load data
qp1 <- read.csv("qp1_data.csv")

```


Animated visuals
=======================================

<div style="float: left; width: 95%;">

+ How many of you have seen this famous [gapminder video of Hans Rosling](https://www.bing.com/videos/search?q=hans+rosling&view=detail&mid=4CC1172C798C8C3BD2A94CC1172C798C8C3BD2A9&FORM=VIRE)

+ Animated visuals can help tell a story

    - especially in a presenation
    - you can tell the story while the visual progresses

+ Animations are especially good for time series

    - can work well for other forms of motion as well
        - for example, how median price changes with number of bedrooms
    - key is that members of a group differ over the varialbe that you are animating over

+ Only use if the animation ads substance

    - do not just use animations to show sizzle

+ You can save these in your TA and call them later

    - `anim_save()` is much like `ggsave()`

</div>


Animated visual basics
=======================================

<div style="float: left; width: 95%;">

+ I want to show you a similar animation to the one Hans used

+ We will be using gganimate

    - I will show several examples of how to use gganminate
    - there are several differnt functions and they continue to develop new ones

+ Here we use the function `transition_time()`

    - this has the visual transition along a variable that is not in the plot
    - see that `year` is not in `ggplot()` but it is in `transition_time(year)`
    
```{r ann1, message = FALSE, warning = FALSE, fig.height = 4, fig.width = 5}
# Load package with gapminder data
library(gapminder)

# Build a faceted animation graph
ggplot(data = gapminder, 
       mapping = aes(x = gdpPercap/1000, y = lifeExp, 
                     size = pop, color = country)) +
  geom_point(alpha = 0.7, show.legend = FALSE) +
  scale_colour_manual(values = country_colors) +
  scale_size(range = c(2, 12)) +
  scale_x_log10() +
  facet_wrap(~continent) +
  theme_bw() +

  # Here comes the gganimate specific bits
  labs(title = "Real GDP and life expectancy grow together: Year {frame_time}",
       x = "GDP per capita ($000)", y = "Life expectancy") +
  transition_time(year) +
  ease_aes("linear")

```

</div>


Time-series animation visuals
=======================================

<div style="float: left; width: 95%;">

+ Here is a typical time series visual

    - here I use `transition_reveal()` by the x-variable in `ggplot()`
    - also set `frame = yr_built` for the visual to transition along

+ I use `annimate()` to set the speed of the reveal

```{r ann2, message = FALSE, warning = FALSE}
# Graph median price by year built
p <- qp1 %>% 
  select(yr_built, waterfront, price) %>% 
  group_by(yr_built, waterfront) %>% 
  summarize(med_price = median(price/1000)) %>% 
  ggplot(aes(x = yr_built, y = med_price, frame = yr_built)) + 
    geom_line(aes(color = as.factor(waterfront))) + 
    scale_color_manual(name = "Waterfront", labels = c("No", "Yes"), values = c("navy", "blue"),
                       guide = guide_legend(reverse=TRUE)) +
    theme_classic() +
    theme(axis.title.x = element_blank(),
          legend.position="bottom") +
  labs(title = "Waterfront median price is much more variable than non-waterfront",
       x = "", y = "Median price ($000)") +  
  transition_reveal(yr_built) 

# Run as a GIF annimation and control speed
animate(p, nframes = 300)

```

</div>

Animating distinct stages of a visual
=======================================

<div style="float: left; width: 95%;">

+ Can also animate visuals that have discrete states

    - good for bar graphs or similar

+ the variable in `transition_state()` does not have to be called in `ggplot()`

    - i.e. could have done by `view`

```{r ann3, message = FALSE, warning = FALSE, fig.height = 4, fig.width = 5}
# Animate a bar graph
qp1 %>% 
  select(price, bedrooms, waterfront) %>% 
  group_by(bedrooms, waterfront) %>%
  filter(bedrooms < 7 & bedrooms > 0) %>% 
  summarise(med_price = median(price/1000)) %>% 
  ggplot(aes(x = bedrooms, y = med_price, fill = as.factor(waterfront))) +
  geom_bar(stat = "identity", position = "dodge") +
  theme_classic() +
  scale_fill_manual(name = "Waterfront", labels = c("No", "Yes"), values = c("lightblue", "blue"),
                    guide = guide_legend(reverse=TRUE)) +
  labs(title = "Median price grows more per bedroom for waterfront homes",
       x = "Number of bedrooms", y = "Median price ($000)") +
  transition_states(bedrooms, wrap = FALSE) +
  shadow_mark()

```

</div>


