---
title: "Lecture 18: Dashboards in R Part 2 - Shiny"
author: "Nick Huntington-Klein"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:   
  revealjs::revealjs_presentation:
    theme: simple
    transition: slide
    self_contained: true
    smart: true
    fig_caption: true
    reveal_options:
      slideNumber: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(tidyverse)
options("kableExtra.html.bsTable" = T)
rinlinevarname <- function(code){
  html <- '<code  class="r">``` `CODE` ```</code>'
  sub("CODE", code, html)
}
```


## Dashboards in R

```{r, results = 'asis'}
cat("
<style>
.reveal section img {
  border: none !important;
  box-shadow: none;
}
</style>")
```

- Dashboards just to show a few graphs is one thing
- And so far what we know how to do is put our own, nicely stylized and customized **ggplot2** graphs on an R dashboard
- Or use something prepackaged that works with **htmlwidgets**
- But what else can we do?

## Interactivity

- Interactivity is a common and expected feature of dashboards
- **plotly** is the easiest way to do this. Of course, by default, the only interaction you get is things like the tooltip, zooming in, etc.
- How can you let people play with your data, change settings?
- How can you make your dashboard an *app*?
- You *could* learn **plotly** fully. It's completely versatile. However, in my experience, it's awful to try to use beyond the defaults
- so....

## Shiny

- Shiny is an R package (and framework) that easily incorporates interactivity into any sort of data presentation you like
- It's highly flexible, and in a pretty easy way, that mostly just lets you do what you already know how to do, then wrap it in a larger system that adds some easy-to-use interactivity
- The way it does this is by splitting the interactivity into a sidebar, which then stores its settings, and can be accessed in other parts of the code.
- Hosting can be wherever, but [shinyapps.io](shinyapps.io) is designed for it and has a free version, and you can upload to it directly and easily with the "Publish" button


## Shiny

- Shiny works by separating the work into three main parts:
- A `global` code chunk that does all the shared activity, like data loading and preprocessing, so it doesn't need to be redone every time someone changes an option
- A **flexdashboard** Column with the `{.sidebar}` attribute with a series of `Input()` functions so the user can select and change inputs and options
- Then the rest of our **flexdashboard** as normal, accessing the elements from the sidebar with `input$`, and instructing **flexdashboard** to update the visualizations with changes using `render()` functions

## Example

- Let's take a look at an example Shiny dashboard and think about how the different features must have been put together
- [https://econometricsbysimulation.shinyapps.io/OLS-App/](https://econometricsbysimulation.shinyapps.io/OLS-App/)

## Shiny and flexdashboard

- Let's walk through [https://rmarkdown.rstudio.com/flexdashboard/shiny.html](https://rmarkdown.rstudio.com/flexdashboard/shiny.html)
- Pay close attention to the source code for the example
- How do the input functions come together, and how are the choices referred to in the visualization?
- (Even deeper documentation at [https://mastering-shiny.org](https://mastering-shiny.org) or the [Shiny Dev Center](https://shiny.rstudio.com/)

## Inputs

- What kinds of inputs can we have?
- Dropdown menu (`selectInput`), a slider (`sliderInput`), radio buttons (`radioButtons`), text (`textInput`), numbers (`numericInput`), a checkbox (`checkboxInput`), dates or date ranges (`dateInput` and `dateRangeInput`) and file upload (`fileInput`)
- For each, there are options for layout and the available choices.
- Let's look at one in more detail, `selectInput`

## selectInput

- This could be well used to, for example, choose which of many ways to look at the data, or which variable to graph, or which subset of the data to look at
- The syntax is `selectInput(inputID, label, choices, selected = NULL, multiple = FALSE, selectize = TRUE, width = NULL, size = NULL)`
- (the = parts are just the defaults)
- The `inputID` is the slot in `input$` where the result will be stored. so with `inputID = 'subset'`, you can later use `input$subset` to know what was selected. That's not what the user sees for a title though, they see `label`

## selectInput

- `choices` are the options, with default `selected`, in a standard vector format. So maybe to choose whether to graph independent or chain restaurants, `choices = c('Choose Restaurant Type' = '','Independent','Chain')`
- `multiple` determines whether multiple options can be selected
- Then the minute details, with presentation on the page with `width`, number to show in the box before scrolling with `size`, and whether and how to use the `selectize` JavaScript package with `selectize` and `options`.

## Output

- You can include output to be updated as the options change with `renderPlot()` (for plots), `renderPrint()` (for any object being printed / shown on its own), `renderTable()` for tables of data, and `renderText()` for actual text output.
- Inside of these functions we just take a regular plot / object / data table / text, refer to elements from the sidebar with `input$inputID`, wrap that in `{}`, and wrap THAT in the appropriate `render` function

## Output

(Using fake data) in the global chunk: `library(tidyverse)` and `data(RestaurantData)`

In the sidebar column, an R chunk with:

```{r, echo = TRUE, eval = FALSE}
selectInput('restaurantType', label = 'Type of Restaurant', choices = choices = c('Choose Restaurant Type' = '','Independent','Chain'))
```

And then in the next column,

```{r, echo = TRUE, eval = FALSE}
renderTable({
  RestaurantType %>% filter(type == input$restaurantType)
})
```

This will show a table of all the data, letting you pick whether to show independent or chain restaurants.

## Example

- Let's see the example dashboard I have here
- I have used the `mtcars` data we've used many times before
- I've included a `geom_bar(stat='summary',fun = 'mean')` graph summarizing the mean of a variable by the type of transmission
- But I let you pick the variable! (with `selectInput`) (Hint: use `aes_string` to input a string variable to **ggplot2**, this means everything must be a string in it)
- And also the color scheme (with `textInput`). Note if you put in something that's not a color (which I check with `%in% colors()`) it will correct to black.

## Dashboard Practice

- Replicate my dashboard
- Then let's see how I did it
- Add one new visualization/table/etc.
- And one new input to control it