---
title: "Lecture 9: Structuring ggplot2"
author: "Nick Huntington-Klein"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:   
  revealjs::revealjs_presentation:
    theme: simple
    transition: slide
    self_contained: true
    smart: true
    fig_caption: true
    reveal_options:
      slideNumber: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(tidyverse)
library(ggrepel)
library(scales)
```



## Structuring ggplot

```{r, results = 'asis'}
cat("
<style>
.reveal section img {
  border: none !important;
  box-shadow: none;
}
</style>")
```

- Last time we went through the basics of the grammar of graphics
- And how to put together a ggplot graph:
- Data, aesthetics, geometry

## Today!

- We'll talk about different ways of changing the *structure* of your ggplot
- Scales and coordinates, labels and legends, new geometries
- There are infinite possibilities here, we can't cover everything
- The goal is to reduce unknown unknowns so you know where to look/Google
- Also, just start typing in RStudio and see what autocompletes!

## The Key Pieces

A common structure for a `ggplot()` command with a fair amount of customization might be:

```{r, echo = TRUE, eval = FALSE}
ggplot(data, aes(x = xvar, y = yvar, color = colorvar)) + 
  geom_mygeometry(otheraesthetic = value) + 
  scale_aesthetic_type(labels = something) + 
  labs(x = 'My X Var', y = 'My Y Var', title = 'My Title') + 
  guides(color = 'none')
```

## Scales

- Every axis (aesthetic element) has a *scale*
- This determines how *a value in the data* (1, 2, 3) turns into *a value on the graph* (100 pixels to the right, 200 pixels, 300 pixels; or "red", "blue", "green")
- These scales are fully manipulable and label-able!
- `scale_axisname_type`

## Discrete and Continuous Scales

- For example:
- `scale_x_continuous` for a continuous x-axis
- `scale_x_discrete` for a discrete one
- Or similarly `scale_y_continuous`, `scale_color_continuous`, `scale_color_gradient`, `scale_fill_discrete`, and so on and so on

## Discrete and Continuous Scales

- Use to:
- Set colors, set limits (want the axis to extend beyond its values? `limits`)
- Name the scale
- Label values (for dicrete values, or perhaps as percents?)

## Setting Values Manually

- `scale_x_manual` for discrete plots will let you set things by hand
- For example if `x` is different companies and you want to pick each company's brand color for its bar

## Examples

```{r, echo = TRUE, fig.height = 4, fig.width = 6}
data <- tibble(category = c('Apple','Banana','Carrot','Apple','Banana','Carrot'),
               person = c('Me','Me','Me','You','You','You'),
               quality = c(.06,.04,.03,.01,.06,.03))
ggplot(data, aes(x = person, y = quality, fill = category)) + geom_col(position = 'dodge')
```

## Setting Scales

```{r, echo = TRUE, fig.height = 4, fig.width = 6}
library(scales)
ggplot(data, aes(x = person, y = quality, fill = category)) + geom_col(position = 'dodge') + 
  scale_y_continuous(labels = label_percent(), limits = c(0,.1)) +
  scale_x_discrete(position = 'top') + 
  scale_fill_manual(values = c('Apple'='red','Banana'='yellow','Carrot'='orange'))
```

## Setting Scales

- `limits` in the `scale` function can be used to specify the vector of values to be included, or as a range for continuous variables
- `labels` is also handy. `labels = c('Red Apple',` `'Yellow Banana','Orange Carrot')` would relabel the legend (or axis labels)
-  In a continuous function, say for dollars, `scale_x_continuous(labels = scales::label_dollar())` would put it in dollar terms. More on this in a moment
- Or dates: `scale_x_date(labels = function(x)` `paste(month.abb[month(x)],year(x)))`

## Setting Color or Fill Scales

- Selecting a color scale is important - we've discussed discrete and continuous palettes
- There are a bunch of `scale_color_`/`scale_fill_` functions that solely exist to help with this!
- (not to mention entire packages like **paletteer**)

## Setting Color or Fill Scales

Especially useful are:

- `scale_color_gradient()` for gradient scales (or `_gradient2()` for diverging scales with a "middle" in them), `scale_color_viridis()` also has some great gradient scales (either discrete or continuous!)
- `scale_color_brewer()`/`scale_fill_brewer()` functions for discrete values, or `_distiller()` for continuous values, or `_fermenter()` for binned

## Setting Color or Fill Scales

```{r, echo = TRUE, fig.height = 4, fig.width = 6}
ggplot(data, aes(x = person, y = quality, fill = category)) + geom_col(position = 'dodge') + 
  scale_y_continuous(labels = label_percent(), limits = c(0,.1)) +
  scale_x_discrete(position = 'top') + 
  scale_fill_brewer(palette = 'Dark2')
```


## Setting Color or Fill Scales

```{r, echo = TRUE, fig.height = 4, fig.width = 6}
ggplot(data, aes(x = person, y = quality, group = category, fill = quality)) + geom_col(position = 'dodge') + 
  scale_y_continuous(labels = label_percent(), limits = c(0,.1)) +
  scale_x_discrete(position = 'top') + 
  scale_fill_viridis_c()
```

## Setting Color or Fill Scales

```{r, echo = TRUE, fig.height = 4, fig.width = 6}
ggplot(data, aes(x = person, y = quality, group = category, fill = quality)) + geom_col(position = 'dodge') + 
  scale_y_continuous(labels = label_percent(), limits = c(0,.1)) +
  scale_x_discrete(position = 'top') + 
  scale_fill_gradient2(midpoint = .03)
```

## Transformations

- We can transform values as we plot them
- Many `scale_something_continuous` entries have a `trans` option, set to `date`, `log`, `probability`, `reciprocal`, `sqrt`, `reverse`, etc. etc. to perform that transformation before plotting 
- You can also do things like `scale_x_log10` or `scale_y_reverse` for some common transforms
- `scale_something_binned()` takes a continuous value and puts it in bins, handy sometimes for simplification

## Transformations

```{r, echo = TRUE, fig.height = 4, fig.width = 6}
ggplot(mtcars, aes(x = mpg, y = hp, color = wt)) + 
  geom_point()
```

## Transformations

```{r, echo = TRUE, fig.height = 4, fig.width = 6}
ggplot(mtcars, aes(x = mpg, y = hp, color = wt)) + 
  geom_point() + 
  scale_x_log10() + 
  scale_y_continuous(trans='reverse') + 
  scale_color_binned()
```

## Log Scales

When to use log scales?

- When data is highly skewed, so a few huge observations are drawing all focus
- When the relationship is multiplicative so you want to show, say, percentage growth

## Labeling Scales for Continuous Variables

Scale functions have a `labels=` option. 

- USE IT!!
- This is the quickest and easiest way to make your graph look a zillion times better and more professional
- We already covered this for categorical variables. `labels = c('A','B')` gives the categories the names A and B, in that order (tip: check the order first)
- For continuous variables, the **scales** package helps us immensely here

## scales

Two main types of functions in **scales**:

- Transformation functions like `dollar()`: `dollar(10)` creates $10 (NOTE: handy sometimes in RMarkdown text! Also note this creates text, not numbers, so don't use them in `aes()` unless you want the variable to be a string)
- Labeling functions like `label_dollar()` designed to slot directly into the `labels=` argument. `scale_y_continuous(labels = label_dollar())` turns all your y-axis labels into the dollar equivalent


## scales

```{r, echo = TRUE, fig.height = 4, fig.width = 6}
ggplot(data, aes(x = person, y = quality, fill = category)) + geom_col(position = 'dodge') + 
  scale_y_continuous(labels = label_percent(), limits = c(0,.1))
```

## Scales

The `label_` functions have lots of options! You can set the `accuracy` (precision), decide how to break up big numbers (`big.mark`) or scale things down to, say, thousands! (`scale=1/1000, suffix = 'k'`)

```{r, echo = TRUE, fig.height = 2, fig.width = 6}
data(gapminder, package = 'gapminder')
ggplot(gapminder, aes(x = gdpPercap, y = lifeExp)) + geom_point() + 
  scale_x_log10(labels = label_dollar(accuracy = 1, scale = 1/1000, suffix = 'k'))
```

## Documentation

- There are a zillion options
- Be sure to use `help(whatever)` before using something
- Not sure what the function is? Check the **ggplot2** documentation page
- Or just start typing `scales_` or whatever and see what pops up in autocomplete in RStudio

## Labels with labs()

- Please label your axes and graphs!!
- Label legends too by naming their axes
- Captions and subcaptions available too
- Annotation and labeling data points we'll do next time

## All Kinds of Labels

```{r, echo = TRUE, fig.height = 4, fig.width = 6}
data(mtcars)
mtcars <- mtcars %>% mutate(CarName = row.names(mtcars))
ggplot(mtcars, aes(x = mpg, y = hp, color = wt)) + geom_point() +
  scale_x_log10() + scale_y_reverse() + scale_color_binned() + 
  labs(x = 'Miles per Gallon', y = 'Horsepower', color = 'Car Weight',
       title = 'Title', subtitle = 'Subtitle', caption = 'Caption')
```

## Moving the Legend

- Working with **ggplot2** legends is one of the trickier things
- we'll do a lot more with aesthetic details next time but for now let's just move it and outline it.

## Moving the Legend

```{r, echo = TRUE, fig.height = 4, fig.width = 6}
ggplot(mtcars, aes(x = mpg, y = hp, color = wt)) + geom_point() +
  scale_x_log10() + scale_y_reverse() + scale_color_binned() + 
  labs(x = 'Miles per Gallon', y = 'Horsepower', color = 'Car Weight',
       title = 'Title', subtitle = 'Subtitle', caption = 'Caption') +
  theme(legend.position = c(.9,.3), legend.box.background = element_rect(color='black'))
```

## Or Removing It


```{r, echo = TRUE, fig.height = 4, fig.width = 6}
ggplot(mtcars, aes(x = mpg, y = hp, color = wt)) + geom_point() +
  scale_x_log10() + scale_y_reverse() + scale_color_binned() + 
  labs(x = 'Miles per Gallon', y = 'Horsepower', color = 'Car Weight') +
  guides(color = 'none')
```

## Overlaid Graphs

- In addition to adding chart elements, we can add whole new geometries over the top
- Some stack naturally (adding a `geom_smooth` best fit line over top, for example)
- For others we may have to change the data or aesthetic as we go

## Overlaid Graphs

```{r, echo = TRUE, fig.height = 4, fig.width = 6}
ggplot(mtcars, aes(x = mpg, y = hp, color = wt)) + geom_point() +
  scale_x_log10() + scale_y_reverse() + scale_color_binned() + 
  labs(x = 'Miles per Gallon', y = 'Horsepower', color = 'Car Weight') +
  geom_smooth(method='lm', se = FALSE)
```

## Overlaid Graphs

```{r, echo = TRUE, fig.height = 4, fig.width = 6}
ggplot(mtcars, aes(x = mpg, y = hp, color = wt)) + geom_point() +
  scale_x_log10() + scale_y_reverse() + scale_color_binned() + 
  labs(x = 'Miles per Gallon', y = 'Horsepower', color = 'Car Weight') +
  geom_text_repel(data = mtcars %>% slice(1:5),aes(label = CarName),hjust=-1)
```

## Facets

- Separate graphs by group!

```{r, echo = TRUE, fig.height = 4, fig.width = 6}
ggplot(mtcars, aes(x = mpg, y = hp)) + geom_point() + 
  facet_wrap('cyl') + 
  labs(x = 'Miles per Gallon', y = 'Horsepower', title = 'Horsepower vs. MPG by Cylinders')
```


## ggforce's facet_zoom

```{r, echo = TRUE, fig.height = 4, fig.width = 6}
library(ggforce)
ggplot(iris, aes(Petal.Length, Petal.Width, colour = Species)) +
  geom_point() +
  facet_zoom(x = Species == 'versicolor')
```

## Multiple-Graph Notes

- For overlaying the *same type* of graph, say two line graphs for two variables, don't `+` different geometries, you're almost always better off doing a `pivot_longer()` first and just using `aes()`.
- For sticking together multiple completely disparate graphs, don't use facets, instead load **patchwork** and combine them as desired - we'll discuss this later

## Imitation and Flattery

- Pick one of these graphs or from the previous lecture
- Go into the geometry's help file and recreate an example, or the graph in these slides
- Mess with the scales and labels! Try to get it to work!
