---
title: "Lecture 6: Exploratory Data Analysis"
author: "Nick Huntington-Klein"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:   
  revealjs::revealjs_presentation:
    theme: simple
    transition: slide
    self_contained: true
    smart: true
    fig_caption: true
    reveal_options:
      slideNumber: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
{
library(tidyverse)
library(lubridate)
library(gghighlight)
library(ggthemes)
library(directlabels)
library(RColorBrewer)
}
rinline <- function(code) {
  sprintf('``` `r %s` ```', code)
}
```

## The Purpose of Analysis

```{r, results = 'asis'}
cat("
<style>
.reveal section img {
  border: none !important;
  box-shadow: none;
}
</style>")
```

- What are we even *doing* with data?
- We want to see what sorts of stuff is in the data, so that we know what things we *should* be reporting on
- This requires us to *explore* the data
- Even if we come in with a strong idea of what we want to find, learning about what the data looks like is a good idea

## The Purpose of Analysis

We want to:

- Understand our data (read the docs, too!!!)
- Detect mistakes
- Get a sense of what our variables look like
- Figure out what some of the relationships are

## The Purpose of Analysis

Always be on the lookout (in EDA and in your code!)

- Things that are different that shouldn't be different ("why does this one person have a height 8x that of anyone else?")
- Things that are the same that shouldn't be the same ("Why is the mean income for Americans and Canadians exactly preciely the same to the 8th decimal place?")
- Relationships that are very surprising ("Why is age negatively correlated with height?")

## Exploring Data

- Univariate non-graphical
- Univariate graphical
- Multivariate non-graphical
- Multivariate graphical

## Univariate Exploratory Analysis

Looking for:

- What kind of values we have
- Are there massive outliers? Are there values that look *incorrect*?
- What is the distribution? Is there skew? If it's a factor, does one category dominate?
- You'll want to do this with nearly every variable you have
- Getting all the little presentation details is less important, but the graph should still be readable (labels etc.) On that note, EDA isn't always *concise* - some tables/graphs will slip off the edge of these slides.

## Univariate Exploratory Analysis

- Think about *what features of the variable it makes sense to explore*
- Central tendencies (mean, median, etc.?)
- Tails? Skew? Percentiles?
- If it's a factor, is showing *all* the categories informative? Do you need to collapse first? Or summarize?

## Univariate Exploratory Analysis Tools

- `summary()` shows lots of good info about a variable's distribution (also works to summarize other objects)

```{r, echo = TRUE}
summary(iris$Sepal.Length)
summary(iris$Species)
```


## Univariate Exploratory Analysis Tools

- `str()` shows variable classes (important!) and some values

```{r, echo = TRUE}
str(iris)
```

## Univariate Exploratory Analysis Tools

- `vtable()` in **vtable** is a more readable and flexible version of that. `lush = TRUE` for more info

```{r, echo = TRUE}
library(vtable)
vtable(iris, lush = TRUE)
```

## Univariate Exploratory Analysis Tools

- `sumtable()` in **vtable** is a full table of summary statistics

```{r, echo = TRUE}
sumtable(iris)
```


## Univariate Exploratory Analysis Tools

- Moving on to graphical tools! In **ggplot**, `geom_density()` and `geom_histogram()` can both show full distributions (when might this *not* be useful?)

```{r, echo = TRUE, fig.height = 3}
ggplot(iris, aes(x = Sepal.Length)) + geom_density()
```

## Univariate Exploratory Analysis Tools

- Boxplots aren't for general-audience consumption but they show some easy summary statistics

```{r, echo = TRUE, fig.height = 3}
ggplot(iris, aes(x = Sepal.Length)) + geom_boxplot() + coord_flip()
```

## Univariate Exploratory Analysis

- Load the `gov_transfers` data from the **causaldata** package
- Read the docs! `help(gov_transfers)`
- Explore the variables that are in there on a univariate basis

## Multivariate Exploratory Analysis

- We want to know *how variables are related* in multivariate analysis
- Again, *fit the analysis to the context*. Don't treat a continuous variable as a factor, and if you treat a discrete variable as continuous know that a scatterplot won't work! etc.
- Don't just fire and forget. Think about whether the output is actually informative.
- more than 2 variables are possible too! Just add more groups/aesthetics.

## Multivariate Exploratory Analysis Tools

- For continuous vs. continuous, nothing wrong with starting with a linear correlation!
- Bivariate OLS is just a rescaled correlation

```{r, echo = TRUE}
cor(iris$Sepal.Length, iris$Sepal.Width)
lm(Sepal.Length ~ Sepal.Width, data = iris)
```


## Multivariate Exploratory Analysis Tools

- If one variable is discrete or a factor, don't underestimate plain ol' `group_by() %>% summarize()`
- In fact, many headaches of trying to get R to do something for you can be avoided by just doing `group_by() %>% summarize()` yourself
- Don't forget `na.rm = TRUE` as appropriate

```{r, echo = TRUE}
iris %>%
  group_by(Species) %>%
  summarise(mean.SL = mean(Sepal.Length, na.rm = TRUE),
            sd.SL = sd(Sepal.Length, na.rm = TRUE))
```

## Multivariate Exploratory Analysis Tools

- `sumtable` has a `group` option for the same purpose (note `summ` can customize the summary functions)

```{r, echo = TRUE}
iris %>% sumtable(group = 'Species')
```

## Multivariate Exploratory Analysis Tools

- `tabyl` in **janitor** is great for discrete vs. discrete; use `adorn` functions to get percents instead of counts. Watch the denominator! It affects interpretation a LOT

```{r, echo = TRUE}
library(janitor)
mtcars %>% tabyl(am, vs) %>%
  adorn_percentages(denominator = 'row') %>%
  adorn_pct_formatting() %>% adorn_title()
```

## Multivariate Exploratory Analysis

- Moving on to graphs!
- If one variable is a factor/discrete you can do pretty much any kind of graph you like, but grouped

```{r, echo = TRUE, fig.height = 3}
mtcars %>% group_by(vs, am) %>%
  summarize(mean_mpg = mean(mpg)) %>%
  ggplot(aes(x = vs, fill = factor(am), y = mean_mpg)) + geom_col(position = 'dodge')
```

## Multivariate Exploratory Analysis

- When it comes to densities, typical geoms look cluttered. I recommend `geom_density_ridges` in **ggridges** or `geom_violin` (or use facets)

```{r, echo = TRUE, fig.height = 3}
library(ggridges); library(patchwork)
p1 <- ggplot(iris, aes(x = Sepal.Length, y = Species, fill = Species)) + geom_density_ridges()
p2 <- ggplot(iris, aes(x = Species, y = Sepal.Length, fill = Species)) + geom_violin()
p1 + p2
```

## Multivariate Exploratory Analysis

- For two continuous variables, things can be tricky, but a scatterplot is a good place to start

```{r, echo = TRUE, fig.height = 4}
ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width)) + geom_point()
```

## Automation

- As a starting place, `ggpairs` in **GGally** will automatically pick univariate and multivariate comparisons to start

```{r, echo = TRUE, eval = FALSE}
GGally::ggpairs(iris)
```

## Automation

```{r}
GGally::ggpairs(iris)
```

## Multivariate Exploratory Analysis

- Let's do some of this in our `gov_transfers` data!

## Going into Detail

- This has all been running with little input from us at this point
- Yes we have to figure out which kind makes sense but we're just exploring
- Going for a bit more detail requires us to  have a *question* to answer, and then we can look into that
- This targets us towards the variables/relationships to study, and *how* to explore them further
- Perhaps, for example, checking trivariate relationships, one example being "*in this subgroup* how are X and Y related?"

## Careful!

Be aware of what EDA does not do:

- Checking *every* relationship will tend to uncover non-real relationships by random chance
- Consider doing EDA on a random subset data for this reason, so you can check if something interesting actually is there in the other/full data and don't trick yourself
- Hardly ever establishes a *causal relationship* - avoid causal language unless you can really back it up!

## Let's Do It (if there's time)

- [https://vincentarelbundock.github.io/Rdatasets/csv/Stat2Data/Clothing.csv](https://vincentarelbundock.github.io/Rdatasets/csv/Stat2Data/Clothing.csv)
- Data description: [https://rdrr.io/rforge/Stat2Data/man/Clothing.html](https://rdrr.io/rforge/Stat2Data/man/Clothing.html)
- Explore! What do we find?