---
title: "Lecture 4: R and RMarkdown"
author: "Nick Huntington-Klein"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:   
  revealjs::revealjs_presentation:
    theme: simple
    transition: slide
    self_contained: true
    smart: true
    fig_caption: true
    reveal_options:
      slideNumber: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
{
library(tidyverse)
library(lubridate)
library(gghighlight)
library(ggthemes)
library(directlabels)
library(RColorBrewer)
}
rinline <- function(code) {
  sprintf('``` `r %s` ```', code)
}
```

## Tech!

```{r, results = 'asis'}
cat("
<style>
.reveal section img {
  border: none !important;
  box-shadow: none;
}
</style>")
```

- As we go through the course, we're going to need to kep data visualization and communication principles close at hand
- But it's also unavoidable that actually using them will require some technical skill on our part
- In this class we'll be using R
- Specifically, the **tidyverse** set of packages, and RMarkdown
- I don't expect you to become R experts. You're not learning R so much as learning **dplyr** and **ggplot2** and some extra details

## R

- Why R instead of Python?
- It's good to know more than one language (nothing ever lasts!)
- **ggplot2** is a top-tier visualization package and makes it much easier to implement our principles
- Data-cleaning is a bit more intuitive than in **pandas** (and faster too if you're using **data.table** but we aren't)
- Python > R in designing software for production, or in some machine learning apps. But we aren't doing those!

## R - The Basics

- Let's switch over to an RStudio walkthrough. We'll cover:
- Getting set up
- The RStudio window
- Using basic R elements
- I'll also show you how to set up the **swirl** package so you can do some walkthroughs on your own

## RMarkdown

- RMarkdown is a text and code processing system
- Text and code in -> Formatted output out (HTML, Word, PDF, slides, books, etc.)
- These slides are made in RMarkdown. I write books and papers in it too.

## Markdown

- Markdown is a "markup language" like HTML or LaTeX. All your formatting goes in as raw text, and then you run it through an interpreter to get output
- Markdown is *super super simple*
- Let's walk through the [Markdown cheatsheet](https://github.com/adam-p/markdown-here/wiki/Markdown-Cheatsheet)
- Note HTML code also works!

## Markdown

- Why Markdown instead of, like, Word? or JuPyteR?
- Can be automated. Includes code and output directly in the document, including for in-line numbers
- Much easier to share via something like, e.g., GitHub
- Output is in standard formats like Word or HTML so they don't need any sort of setup to see it

## The YAML

- The YAML is the set of text at the top of an RMarkdown that tells it how to interpret all that text. Here's the YAML for these slides:

```{r, eval = FALSE, echo = TRUE}
---
title: "Lecture 4: R and RMarkdown"
author: "Nick Huntington-Klein"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:   
  revealjs::revealjs_presentation:
    theme: simple
    transition: slide
    self_contained: true
    smart: true
    fig_caption: true
    reveal_options:
      slideNumber: true
---
```

## The YAML

- Here's a more standard YAML, the default from File -> New Rmarkdown:
- Common option: toc: TRUE to turn on the table of contents on the left

```{r, eval = FALSE, echo = TRUE}
---
title: "Untitled"
author: "Nick Huntington-Klein"
date: "7/5/2021"
output: html_document
---
```

## Code Chunks

- Include code in-line (fantastic for memos) with `r rinline('mean(1:3)')` which makes `r mean(1:3)`
- Include code chunks with triple backticks. Let's take a look at an example RMarkdown doc
- Note: want a data frame to nicely format as a table? use `knitr::kable()`.

## Code Chunks

- Common options to set in each code chunk, or altogther in `knitr::opts_chunk$set()`:
- `echo`: Show code or not?
- `eval`: Evaluate the code inside or not?
- `warning` and `message`: Show warnings and messages from the code?
- `include`: Include *any output at all* or not?

## Let's do it

- Make a new RMarkdown HTML document
- YAML: Turn the toc on, title it and add your name
- Make one big section and title it (#) and two small (##)
- In the setup chunk, load the **tidyverse** and load `data(storms)`
- Write some text in the first small one, including a link to something
- In the second, add a code chunk that does a `summary()` of `storms`
- Then, some more text, with an in-line text to get the mean of `storms$wind`.
- Bonus: load the **scales** package and use `number()` to format the in-line text.