---
title: "Lecture 8: Styling ggplot2"
author: "Nick Huntington-Klein"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:   
  revealjs::revealjs_presentation:
    theme: simple
    transition: slide
    self_contained: true
    smart: true
    fig_caption: true
    reveal_options:
      slideNumber: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(tidyverse)
library(ggrepel)
```



## Styling ggplot

```{r, results = 'asis'}
cat("
<style>
.reveal section img {
  border: none !important;
  box-shadow: none;
}
</style>")
```

- We should have a decent idea at this point of how to put a **ggplot2** graph together
- But how do we make it look good?

## Today!

- We'll talk about how to set aesthetic attributes along every axis
- As well as overall styling and theming
- Highlighting, putting graphs together, making them interactive, annotating...

## Aesthetic Characteristics

- As you'd expect, we'll be setting the values of certain aesthetic properties
- Which ones apply depend on the geometry. See [this page](https://ggplot2.tidyverse.org/articles/ggplot2-specs.html). Some common ones:
- color, linetype, fill, size, shape
- Text aesthetics: family (font), fontface, hjust, vjust (see **extrafont** package to use all your system fonts)

## Decorating Axes vs. Decorating Geometries

- When a characteristic is set *inside* of an `aes()`, it becomes an axis and must be mapped to a variable name
- *Outside* of an `aes()` (and in the geometry), it's a setting applied to the entire geometry

## Aesthetic Characteristics

```{r, echo = TRUE, fig.height = 4, fig.width = 6}
# We've seen this before
mtcars <- mtcars %>%
  mutate(Transmission = factor(am, labels = c('Automatic','Manual')))
ggplot(mtcars, aes(x = mpg, y = hp, color = Transmission)) + 
  geom_point()
```

## Aesthetic Characteristics

```{r, echo = TRUE, fig.height = 4, fig.width = 6}
ggplot(mtcars, aes(x = mpg, y = hp, color = Transmission, 
                   size = wt, shape = Transmission)) + 
  geom_point()
```

## Aesthetic Characteristics

```{r, echo = TRUE, fig.height = 4, fig.width = 6}
ggplot(mtcars, aes(x = mpg, y = hp, color = Transmission, 
                   size = wt)) + 
  geom_point(shape = 10)
```

## Aesthetic Characteristics

```{r, echo = TRUE, fig.height = 4, fig.width = 6}
ggplot(mtcars, aes(x = Transmission, fill = factor(cyl))) + 
  geom_bar(position = 'dodge', linetype = 'dashed', color = 'black')
```

## Themes

- Everything *behind* the geometries can be manipulated as well
- For example, we can change the coordinate system (usually to `coord_flip` it for bar graphs)
- Or, we can change the legends, axis labels, etc., with `theme()`.
- The options are endless! See `help(theme)`

## Themes

```{r, echo = TRUE, fig.height = 4, fig.width = 6}
library(ggalt)
mtcars2 <- mtcars %>% group_by(Transmission) %>% summarize(count = n())
ggplot(mtcars2, aes(x = Transmission,y=count)) + geom_lollipop()
```

## Themes

```{r, echo = TRUE, fig.height = 4, fig.width = 6}
# geom_lollipop in particular has a 'horizontal' option anyway
# so this is just for demonstration
ggplot(mtcars2, aes(x = Transmission,y=count)) + 
  geom_lollipop(color = 'red', size = 2) + coord_flip()
```

## Axes

- We can control axis text with `axis.text` and the ticks we mark on the `scale`
- `element_text` provides text objects

```{r, echo = TRUE, eval = FALSE, fig.height = 4, fig.width = 6}
ggplot(mtcars2, aes(x = Transmission,y=count)) + 
  geom_lollipop(color = 'red', size = 2) + coord_flip() + 
  labs(x = '', y = '') + 
  scale_y_continuous(breaks = c(10,20), limits = c(0,20)) + 
  theme(axis.text.x = element_text(size = 12,family='serif'),
        axis.text.y = element_text(size = 16,family='serif', angle = 10))
```

## Axes

```{r, echo = FALSE, eval = TRUE, fig.height = 4, fig.width = 6}
ggplot(mtcars2, aes(x = Transmission,y=count)) + 
  geom_lollipop(color = 'red', size = 2) + coord_flip() + 
  labs(x = '', y = '') + 
  scale_y_continuous(breaks = c(10,20), limits = c(0,20)) + 
  theme(axis.text.x = element_text(size = 12,family='serif'),
        axis.text.y = element_text(size = 16,family='serif', angle = 10))
```


## Axes

- Order for discrete axes can be changed with `limits`

```{r, echo = TRUE, eval = TRUE, fig.height = 4, fig.width = 6}
ggplot(mtcars2, aes(x = Transmission,y=count)) + 
  geom_lollipop(color = 'red', size = 2) + coord_flip() +
  scale_x_discrete(limits = c("Manual", "Automatic"))
```

## Axes

- Change axis and tick lines with `axis.line` and `element_line`
- Most everything can be changed generally (`axis.line` or even `line`) or specifically (`axis.ticks.x`)

```{r, echo = TRUE, eval = FALSE, fig.height = 4, fig.width = 6}
ggplot(mtcars2, aes(x = Transmission,y=count)) + 
  geom_lollipop(color = 'red', size = 2) + coord_flip() + 
  labs(x = '', y = '') + 
  scale_y_continuous(breaks = c(10,20), limits = c(0,20)) + 
  theme(axis.line = element_line(color = 'red'),
        axis.ticks.x = element_line(size = 5))
```

## Axes

```{r, echo = FALSE, eval = TRUE, fig.height = 4, fig.width = 6}
ggplot(mtcars2, aes(x = Transmission,y=count)) + 
  geom_lollipop(color = 'red', size = 2) + coord_flip() + 
  labs(x = '', y = '') + 
  scale_y_continuous(breaks = c(10,20), limits = c(0,20)) + 
  theme(axis.line = element_line(color = 'red'),
        axis.ticks.x = element_line(size = 5))
```

## Background

- Work on the `panel` to change what goes behind that geometry! `element_rect` might come up!
- Pretty much ANY element can be eliminated with `element_blank()`
- This gonna be ugly

```{r, echo = TRUE, eval = FALSE, fig.height = 4, fig.width = 6}
ggplot(mtcars2, aes(x = Transmission,y=count)) + 
  geom_lollipop(color = 'red', size = 2) + coord_flip() + 
  theme(panel.background = element_rect(color = 'blue', fill = 'yellow'),
        panel.grid.major.x = element_line(color = 'green'),
        panel.grid.major.y = element_blank())
```

## Background

```{r, echo = FALSE, eval = TRUE, fig.height = 4, fig.width = 6}
ggplot(mtcars2, aes(x = Transmission,y=count)) + 
  geom_lollipop(color = 'red', size = 2) + coord_flip() + 
  theme(panel.background = element_rect(color = 'blue', fill = 'yellow'),
        panel.grid.major.x = element_line(color = 'green'),
        panel.grid.major.y = element_blank())
```

## Legends

- We showed a bit last time how legends can be changed.
- Note in `element_text` (and elsewhere, like annotation or `geom_text`), `hjust` and `vjust` align text horizontally and vertically. Set 0/1 for L/R or Top/Bottom
- `legend.position` is in portion of the frame. .8 = 80% to the right.

```{r, echo = TRUE, eval = FALSE, fig.height = 4, fig.width = 6}
ggplot(mtcars, aes(x = mpg, y = hp, color = Transmission)) + geom_point() +
  theme(legend.text = element_text(hjust = .5, family = 'serif'),
        legend.title = element_text(hjust = 1, family = 'serif', face = 'bold'),
        legend.background = element_rect(color = 'black', fill = 'white'),
        legend.position = c(.8,.7))
```

## Legends

```{r, echo = FALSE, eval = TRUE, fig.height = 4, fig.width = 6}
ggplot(mtcars, aes(x = mpg, y = hp, color = Transmission)) + geom_point() +
  theme(legend.text = element_text(hjust = .5, family = 'serif'),
        legend.title = element_text(hjust = 1, family = 'serif', face = 'bold'),
        legend.background = element_rect(color = 'black', fill = 'white'),
        legend.position = c(.8,.7))
```

## Themes

- The theme is immensely flexible. We didn't even cover everything, like the plot title/caption
- You can save your `theme()` settings as an object to use repeatedly as a house style
- Or use a prepackaged one! You can even tack on more `theme()` customization afterwards.
- Many `theme_` settings you can tack on. I only use a few...
- (many many others... want to look like FiveThirtyEight or The Economist? Check **ggthemes**)

## Prepackaged Themes: Default

```{r, echo = TRUE, fig.height = 4, fig.width = 6}
ggplot(mtcars, aes(x = mpg, y = hp, color = Transmission)) + geom_point()
```

## Prepackaged Themes: Minimal (my go-to)

```{r, echo = TRUE, eval = TRUE, fig.height = 4, fig.width = 6}
ggplot(mtcars, aes(x = mpg, y = hp, color = Transmission)) + geom_point() +
  theme_minimal()
```

## Prepackaged Themes: void (for things like flowcharts)

```{r, echo = TRUE, eval = TRUE, fig.height = 4, fig.width = 6}
ggplot(mtcars, aes(x = mpg, y = hp, color = Transmission)) + geom_point() +
  theme_void()
```

## Prepackaged Themes: tufte (in ggthemes)

```{r, echo = TRUE, eval = TRUE, fig.height = 4, fig.width = 6}
library(ggthemes)
ggplot(mtcars, aes(x = mpg, y = hp, color = Transmission)) + geom_point() +
  theme_tufte()
```

## Prepackaged Themes: bbc_style (in bbplot)

```{r, echo = TRUE, eval = TRUE, fig.height = 4, fig.width = 6}
# devtools::install_github('bbc/bbplot')
library(bbplot)
ggplot(mtcars, aes(x = mpg, y = hp, color = Transmission)) + geom_point() +
  bbc_style()
```

## Other Styling

- Let's move on to some other ways we can build focus and clarity
- Highlight and annotations!

## Highlighting

- Highlight certain values and gray others with the **gghighlight** package

```{r, echo = TRUE, eval = FALSE, fig.height = 4, fig.width = 6}
library(gghighlight); data(gapminder, package = 'gapminder')
ggplot(gapminder, aes(x = year, y = lifeExp, color=country)) + geom_line(size = 1.5) + 
  labs(x = NULL, y = "Life Expectancy", title = "North America Only") + 
  scale_x_continuous(limits=c(1950,2015),
                     breaks = c(1950,1970,1990,2010))+
  gghighlight(country %in% c('United States','Canada','Mexico'),
              unhighlighted_params = aes(size=.1), 
              label_params=list(direction='y',nudge_x=10)) + 
  theme_minimal(base_family='serif') 
```

## Highlighting 


```{r, echo = FALSE, eval = TRUE, fig.height = 4, fig.width = 6}
library(gghighlight); data(gapminder, package = 'gapminder')
ggplot(gapminder, aes(x = year, y = lifeExp, color=country)) + geom_line(size = 1.5) + 
  labs(x = NULL, y = "Life Expectancy", title = "North America Only") + 
  scale_x_continuous(limits=c(1950,2015),
                     breaks = c(1950,1970,1990,2010))+
  gghighlight(country %in% c('United States','Canada','Mexico'),
              unhighlighted_params = aes(size=.1), 
              label_params=list(direction='y',nudge_x=10)) + 
  theme_minimal(base_family='serif') 
```

## Annotation

- There are also many options for *annotating* our data
- For direct data point annotation, especially of line graphs, I recommend `geom_dl` from **directlabels**
- For putting a block of text I recommend **ggannotate**
- I also recommend looking into [**ggtext**](https://wilkelab.org/ggtext/) which allows rich-text formatting of all text on the graph, but we won't be covering it here.

## Direct Line Labels

```{r, echo = TRUE, eval = TRUE, fig.height = 4, fig.width = 6}
library(directlabels)
ggplot(gapminder %>% filter(country %in% c('United States','Canada','Mexico')), 
       aes(x = year, y = lifeExp, color=country)) + geom_line(size = 1.5) + 
  scale_x_continuous(limits = c(1950,2025))+ theme_minimal()+guides(color=FALSE)+
  geom_dl(aes(x = year + 2,label = country),method = 'last.bumpup')
```

## Overall Annotation

- You can use the regular **ggplot2** `annotate()` function, but this takes some work
- **ggannotate** will let you work hands-on with it, and it gives you code at the end
- Also works for adding lines and arrows
- Excel-esque but the end result is more rigorous
- Install with `remotes::install_github` `("mattcowgill/ggannotate")`, not `install.packages()`

## ggannotate

- Let's see what happens when we run this...

```{r, echo = FALSE, eval = TRUE}
p <- ggplot(gapminder %>% filter(country %in% c('United States','Canada','Mexico')), 
       aes(x = year, y = lifeExp, color=country)) + geom_line(size = 1.5) + 
  scale_x_continuous(limits = c(1950,2025))+ theme_minimal()+guides(color=FALSE)+
  geom_dl(aes(x = year + 2,label = country),method = 'last.bumpup')
```

```{r, echo = TRUE, eval = FALSE, fig.height = 4, fig.width = 6}
p <- ggplot(gapminder %>% filter(country %in% c('United States','Canada','Mexico')), 
       aes(x = year, y = lifeExp, color=country)) + geom_line(size = 1.5) + 
  scale_x_continuous(limits = c(1950,2025))+ theme_minimal()+guides(color=FALSE)+
  geom_dl(aes(x = year + 2,label = country),method = 'last.bumpup')

ggannotate::ggannotate(p)
```

## ggannotate

- We get code chunks to add to our graph to get our annotations

```{r, echo=FALSE, eval =TRUE, fig.height = 5, fig.width = 7}
p + geom_label(data = data.frame(x = 1990.89294750866,
y = 62.3489741987567, label = "Mexico catches up considerably\nthrough the latter half of the 20th century."), mapping = aes(x = x, y = y, label = label), size = 3.86605783866058, angle = 0L,
lineheight = 1L, hjust = 0L, vjust = 0.5, colour = "black", family = "sans", fontface = "plain",
label.padding = structure(0.25, class = "unit", valid.unit = 3L, unit = "lines"),
label.size = 0.25,
label.r = structure(0.15,
class = "unit",
valid.unit = 3L,
unit = "lines"),
inherit.aes = FALSE,
show.legend = FALSE) + 
  geom_curve(data = data.frame(x = 1990.89294750866,
y = 63.7041493339792,
xend = 1985.22434845418,
yend = 66.7048942762574),
mapping = aes(x = x,
y = y,
xend = xend,
yend = yend),
size = 0.5,
angle = 61L,
colour = "black",
curvature = -0.45,
arrow = structure(list( angle = 30,
length = structure(0.1,
class = "unit",
valid.unit = 2L,
unit = "inches"),
ends = 2L,
type = 2L),
class = "arrow"),
inherit.aes = FALSE,
show.legend = FALSE)

```

## Export

- Alright, we've crafted our perfect graph
- What do we do now?
- `ggsave()` will save our result to file (or we can Export Image in the Plots frame of RStudio)
- But we have some more options still!

## Sticking Graphs Together

- If you have multiple plots that go together, consider using **patchwork** to put them all into one
- With the library loaded, you can just add plots together and they get stitched!
- Let's stitch together some of the plots from this presentation
- `+` to stick together, `|` to put "beside", `-` to move to next column, `/` for the next row


## Patchwork

```{r, echo = FALSE, eval = TRUE}
p1 <- ggplot(mtcars, aes(x = mpg, y = hp, color = Transmission)) + geom_point() +
  theme_tufte()
p2 <- ggplot(mtcars, aes(x = mpg, y = hp, color = Transmission, 
                   size = wt, shape = Transmission)) + 
  geom_point()
p3 <- ggplot(mtcars2, aes(x = Transmission,y=count)) + 
  geom_lollipop(color = 'red', size = 2) + coord_flip() + 
  theme(panel.background = element_rect(color = 'blue', fill = 'yellow'),
        panel.grid.major.x = element_line(color = 'green'),
        panel.grid.major.y = element_blank())
p4 <- ggplot(mtcars, aes(x = Transmission, fill = factor(cyl))) + 
  geom_bar(position = 'dodge', linetype = 'dashed', color = 'black')
```


```{r, echo= TRUE, eval = TRUE, fig.height = 5, fig.width = 12}
library(patchwork)
# Named and saved these earlier
(p1 | p2 - p3)/p4
```

## Plotly

- If you want an interactive graph, an easy way to do it is to use **plotly** to export
- Often it will just sort of... work!
- THere may be need for further customization, see the docs
- Can help make really busy graphs still useful, e.g. [this one](https://nickchk.com/viz/scorecard/)

```{r, echo = TRUE, eval = FALSE}
library(plotly)
ggplotly(p1)
```

---

```{r, echo = FALSE, eval = TRUE}
library(plotly)
ggplotly(p1)
```



## Practice

- We may not have time today to pracice much of this
- But I want you to take a basic graph, a real basic one, like this:
- `ggplot(mtcars, aes(x = mpg, y = hp)) +` `geom_point()`
- And just see how you can make it look. Mess around with it
- The only goal is to try stuff out.