---
title: "Data Wrangling in the Tidyverse"
author: "Nick Huntington-Klein"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:   
  revealjs::revealjs_presentation:
    theme: simple
    transition: slide
    self_contained: true
    smart: true
    fig_caption: true
    reveal_options:
      slideNumber: true
      
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning =FALSE, message = FALSE)
library(tidyverse)
library(DT)
library(purrr)
library(readxl)
library(lubridate)
```

## Data Wrangling

```{r, results = 'asis'}
cat("
<style>
.reveal section img {
  border: none !important;
  box-shadow: none;
}
</style>")
```

- Data wrangling is the process of going from raw data to something usable
- It is what you will spend most of your time as a data person doing, so you better be good at it!
- Is it part of data communications? Yes, although not really any moreso than your other data classes
- But do they cover it enough? Never! So we do here.
- You will thank me later.

## Data Wrangling

This week:

- The goal of data wrangling
- How to think about data wrangling
- Technical tips for data wrangling in R using the **tidyverse** package (which, importantly, contains the **dplyr** and **tidyr** packages inside, as well as **ggplot2** which we'll use for visualization)
- This is adapted from my workshop on data wrangling in the **tidyverse**. [Video from that is here](https://twitter.com/nickchk/status/1350500257654718465), as well as versions with **data.table** (R package with different syntax, very good for big data sets and extra speed) and **pandas** in Python.


## Tech note: the pipe

`scales::percent(mean(mtcars$am,` `na.rm = TRUE), accuracy = .1)` can be rewritten

```{r, eval = FALSE, echo = TRUE}
mtcars %>%
  pull(am) %>%
  mean(na.rm = TRUE) %>%
  scales::percent(accuracy = .1)
```

- Like a conveyer belt! Nice and easy. 
- `pull()` is a **dplyr** function that says "give me back this one variable instead of a data set" but in a pipe-friendly way, so `mtcars %>% pull(am)` is the same as `mtcars$am` or `mtcars[['am']]`



## Limitations

- We only have so much time! I won't be going into *great* detail on the use of all the technical commands, but by the end of this you will know what's out there and generally how it's used
- *As with any computer skill, a teacher's comparative advantage is in letting you know what's out there. The* **real learning** *comes from practice and Googling. So take what you see here today, find yourself a project, and do it! It will be awful but you will learn an astounding amount by the end*

## The Core of Data Wrangling

- Always **look directly at your data so you know what it looks like**
- Always **think about what you want your data to look like when you're done**
- Think about **how you can take information from where it is and put it where you want it to be**
- After every step, **look directly at your data again to make sure it's doing what you think it's doing**

I help a lot of people with their problems with data wrangling. Their issues are almost always *not doing one of these four things*, much more so than having trouble coding or anything like that

## The Core of Data Wrangling

- How can you "look at your data"?
- Literally is one way - click on the data set, or do `View()` to look at it
- Summary statistics tables: `sumtable()` or `vtable(lush = TRUE)` in **vtable** for example
- Checking what values it takes: `unique()` or `summary()` on individual variables
- Check if it takes a certain value ever: `sum(data$variable == value)`
- Look for: What values are there, what the observations look like, presence of missing or unusable data, how the data is structured

## The Stages of Data Wrangling

For the purposes of this class:

- From data to tidy data
- From tidy data to data for your analysis

## What we Can and Can't Cover

- We could be at this for literally weeks and not cover everything
- And you wouldn't remember it anyway
- I will cover the stuff we can use immediately and will use all the time
- The rest will be reducing *unknown unknowns*
- **If you need to do something later, and we didn't cover it, maybe come back and check these slides!!**
- The video series, linked on Canvas and [here](https://www.youtube.com/watch?v=EdPKcy1WKD0&list=PLcTBLulJV_AJ4M_a7oYpURLZ36YvUG78S), will go into these things in more detail when you need them 


# From Data to Tidy Data 

## From Data to Tidy Data

- **Data** is any time you have your records stored in some structured format
- But there are many such structures! They could be across a bunch of different tables, or perhaps a spreadsheet with different variables stored randomly in different areas, or one table per observation
- These structures can be great for *looking up values*. That's why they are often used in business or other settings where you say "I wonder what the value of X is for person/day/etc. N"
- They're rarely good for *doing analysis* (calculating statistics, fitting models, making visualizations)
- For that, we will aim to get ourselves *tidy data*

## Tidy Data

In tidy data:

1. Each variable forms a column
1. Each observation forms a row
1. Each type of observational unit forms a table

```{r}
df <- data.frame(Country = c('Argentina','Belize','China'), TradeImbalance = c(-10, 35.33, 5613.32), PopulationM = c(45.3, .4, 1441.5))
datatable(df)
```

## Tidy Data

The variables in tidy data come in two types:

1. *Identifying Variables*/*Keys* are the columns you'd look at to locate a particular observation. 
1. *Measures*/*Values* are the actual data.

Which are they in this data?

```{r}
df <- data.frame(Person = c('Chidi','Chidi','Eleanor','Eleanor'), Year = c(2017, 2018, 2017, 2018), Points = c(14321,83325, 6351, 63245), ShrimpConsumption = c(0,13, 238, 172))
datatable(df)
```
## Tidy Data

- *Person* and *Year* are our identifying variables. The combination of person and year *uniquely identifies* a row in the data. Our "observation level" is person and year. There's only one row with Person == "Chidi" and Year == 2018
- *Points* and *ShrimpConsumption* are our measures. They are the things we have measured for each of our observations
- Notice how there's one row per observation (combination of Person and Year), and one column per variable
- Also this table contains only variables that are at the Person-Year observation level. Variables at a different level (perhaps things that vary between Person but don't change over Year) would go in a different table, although this last one is less important

## Tidying Non-Tidy Data

- So what might data look like when it's *not* like this, and how can we get it this way? 
- Here's one common example, a *count table* (not tidy!) where each column is a *value*, not a *variable*

```{r}
data("relig_income")
datatable(relig_income)
```

## Tidying Non-tidy Data

- Here's a simple and very common case, where we have a category in each column
- This will mess us up if we, say, want to categorize by a "category" variable!

```{r}
multicat = tibble(Date = ymd('2000-01-01') + months(0:3), Sales_Goods = c(4, 5, 1, 2), Sales_Services = c(4, 9, 2, 3))
datatable(multicat)
```

## Tidying Non-tidy Data

There are many more elements to this, but I will cover two tools: `pivot` (briefly) and `join` (in more detail)

Pivot will be our first **review in a flash here, just remember that it exists, come back later or watch the video when you actually need it**

## Tidying Non-Tidy Data

- The first big tool in our tidying toolbox is the *pivot*
- A pivot takes a single row with K columns and turns it into K rows with 1 column, using the identifying variables/keys to keep things lined up. 
- This can also be referred to as going from "wide" data to "long" data
- Long to wide is also an option
- In every statistics package, pivot functions are notoriously fiddly. Always read the help file, and do trial-and-error! Make sure it worked as intended.

## Pivot

- In the **tidyverse** we have the functions `pivot_longer()` and `pivot_wider()`. Here we want wide-to-long so we use `pivot_longer()`
- This asks for:
- `data` (the data set you're working with, also the first argument so we can pipe to it)
- `cols` (the columns to pivot) - it will assume anything not named here are the keys
- `names_to` (the name of the variable to store which column a given row came from, here "week")
- `values_to` (the name of the vairable to store the value in)
- Many other options (see `help(pivot_longer)`)

## Pivot

```{r, echo = TRUE, eval = TRUE}
# pivot is pretty good at just figuring stuff out for us, at least in this simple case!
multicat_long <- multicat %>%
  pivot_longer(cols = starts_with('Sales_'), # tidyselect functions help us pick columns based on name patterns
               names_to = 'Category', values_to = 'Sales')
multicat_long
```


## Variables Stored as Rows

- Here we have tax form data where each variable is a row, but we have multiple tables For this one we can use `pivot_wider()`, and then combine multiple individuals with `bind_rows()`

```{r}
taxdata <- data.frame(TaxFormRow = c('Person','Income','Deductible','AGI'), Value = c('James Acaster',112341, 24000, 88341))
taxdata2 <- data.frame(TaxFormRow = c('Person','Income','Deductible','AGI'), Value = c('Eddie Izzard',325122, 16000,325122 - 16000))
datatable(taxdata)
```

## Variables Stored as Rows

- `pivot_wider()` needs:
- `data` (first argument, the data we're working with)
- `id_cols` (the columns that give us the key - what should it be here?)
- `names_from` (the column containing what will be the new variable names)
- `values_from` (the column containing the new values)
- Many others! See `help(pivot_wider)`

## Variables Stored as Rows

```{r, echo = TRUE}
taxdata %>%
  pivot_wider(names_from = 'TaxFormRow',
              values_from = 'Value')
```

(note that the variables are all stored as character variables not numbers - that's because the "person" row is a character, which forced the rest to be too. we'll go through how to fix that later)

## Merging Data

- Commonly, you will need to link two datasets together based on some shared keys
- For example, if one dataset has the variables "Person", "Year", and "Income" and the other has "Person" and "Birthplace"

```{r}
person_year_data <- data.frame(Person = c('Ramesh','Ramesh','Whitney', 'Whitney','David','David'), Year = c(2014, 2015, 2014, 2015,2014,2015), Income = c(81314,82155,131292,141262,102452,105133))
person_data <- data.frame(Person = c('Ramesh','Whitney'), Birthplace = c('Crawley','Washington D.C.'))
datatable(person_year_data)
```

## Merging Data

That was `person_year_data`. And now for `person_data`:

```{r}
datatable(person_data)
```

## Merging Data

- The **dplyr** `join` family of functions will do this (see `help(join)`). The different varieties just determine what to do with rows you *don't* find a match for. `left_join()` keeps non-matching rows from the first dataset but not the second, `right_join()` from the second not the first, `full_join()` from both, `inner_join()` from neither, and `anti_join()` JUST keeps non-matches

## Merging Data

```{r, echo = TRUE}
person_year_data %>%
  left_join(person_data, by = 'Person')
```

```{r, echo = TRUE}
person_year_data %>%
  right_join(person_data, by = 'Person')
```

## Merging Data

- Things work great if the list of variables in `by` is the exact observation level in *at least one* of the two data sets
- But if there are multiple observations per combination of `by` variables in *both*, that's a problem! It will create all the potential matches, which may not be what you want:

```{r, echo = TRUE}
a <- tibble(Name = c('A','A','B','C'), Year = c(2014, 2015, 2014, 2014), Value = 1:4)
b <- tibble(Name = c('A','A','B','C','C'), Characteristic = c('Up','Down','Up','Left','Right'))
a %>% left_join(b, by = 'Name')

```

## Merging Data

- This is why it's *super important* to always know the observation level of your data. You can check it by seeing if there are any duplicate rows among what you *think* are your key variables: if we think that `Person` is the key for data set `a`, then `a %>% select(Person) %>% duplicated() %>% max()` will return `TRUE`, showing us we're wrong
- At that point you can figure out how you want to proceed - drop observations so it's the observation level in one? Accept the multi-match? Pick only one of the multi-matches?

## Try It!

Join `data1` with `data2`, maintaining both categories

```{r, echo = TRUE, eval = TRUE}
data1 <- tibble(Month = 1:4, MonthName = c('Jan','Feb','Mar','Apr'),
                Sales = c(6,8,1,2), RD = c(4,1,2,4)) %>%
  pivot_longer(cols = c(Sales, RD), names_to = 'Department')
data2 <- tibble(Department = 'Sales', Director = 'Lavy')
```

## Try It!

```{r, echo = TRUE}
data1 %>%
  left_join(data2) # full_join would also work to retain RD despite no match
```

# From Tidy Data to Your Analysis

## From Tidy Data to Your Analysis

- Okay! We now have, hopefully, a nice tidy data set with one column per variable, one row per observation, we know what the observation level is!
- That doesn't mean our data is ready to go! We likely have plenty of cleaning and manipulation to go before we are ready for analysis
- We will be doing this mostly with **dplyr**
- You'll (a) **definitely want to know this**, and (b) will almost certainly find it easier to do with these **dplyr** tools than with something else you Google

## dplyr

- **dplyr** uses a *small set of "verbs"* to very flexibly do all kinds of data cleaning and manipulation
- The primary verbs are: `filter(), select()`, `arrange()`, `mutate()`, `group_by()`, and `summarize()`.
- Other important functions in **dplyr**: `pull()` (which we covered), `case_when()`, `rename()`, and `slice()`
- See the [dplyr cheatsheet](https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf)

# Subsetting Data

## filter()

- `filter()` limits the data to the observations that fulfill a certain *logical condition*. It *picks rows*.
- For example, `Income > 100000` is `TRUE` for everyone with income above 100000, and `FALSE` otherwise. `filter(data, Income > 100000)` would return just the rows of `data` that have `Income > 100000`

```{r, echo = TRUE}
person_year_data %>%
  left_join(person_data, by = 'Person') %>%
  filter(Income > 100000)
```

## Logical Conditions

- A lot of programming in general is based on writing logical conditions that check whether something is true
- In R, if the condition is true, it returns `TRUE`, which turns into 1 if you do a calculation with it. If false, it returns `FALSE`, which turns into 0. (tip: `ifelse()` is rarely what you want, and `ifelse(condition, TRUE, FALSE)` is redundant)

## Logical Conditions Tips

Handy tools for constructing logical conditions:

`a > b`, `a >= b`, `a < b`, `a <= b`, `a == b`, or `a != b` to compare two numbers and check if `a` is above, above-or-equal, below, below-or-equal, equal (note `==` to check equality, not `=`), or not equal

`a %in% c(b, c, d, e, f)` is SUPER HANDY and checks whether `a` is any of the values `b, c, d, e,` or `f`. Works for text too!

Reverse conditions with `!`, or chain together with `&` (and) or `|` (or)

## slice()

- Like `filter()`, `slice()` picks rows. The difference is that `filter()` picks them by logical condition, while `slice()` does it by row index number
- `slice(1:5)` picks the first five rows, `slice(-103)` drops the 103rd row, etc.


## select()

- `select()` gives you back just a subset of the columns. It *picks columns*
- It can do this by name or by column number
- Use `-` to *not* pick certain columns

If our data has the columns "Person", "Year", and "Income", then all of these do the same thing:

```{r, echo = TRUE}
no_income <- person_year_data %>% select(Person, Year)
no_income <- person_year_data %>% select('Person','Year')
no_income <- person_year_data %>% select(1:2)
no_income <- person_year_data %>% select(-Income)
print(no_income)
```

## arrange()

- `arrange()` sorts the data. That's it! Give it the column names and it will sort the data by those columns.
- Can be handy with `slice()`. `arrange(-Income) %>% slice(1:5)` gives you the five rows with the highest income, for example
- There are also some data manipulation tricks that rely on the position of the data

```{r, echo = TRUE}
person_year_data %>%
  arrange(Person, Year)
```

## Try It!

Load the `construction` data and:

1. `select()` the Year, Month, and West columns
1. `filter()` to just the months of July, August, and September (use `%in%`)
1. `arrange()` by West in descending order (hint: `-West`)
1. `slice()` to pick the top two rows

```{r, echo = TRUE}
data("construction")
```

## Try It!

```{r, echo = TRUE}
construction %>%
  select(Year, Month, West) %>%
  filter(Month %in% c('July','August','September')) %>%
  arrange(-West) %>%
  slice(1:2)
```

# Creating Variables


## mutate()

- `mutate()` *assigns columns/variables*, i.e. you can create variables with it (note also its sibling `transmute()` which does the same thing and then drops any variables you don't explicitly specify in the function)
- You can assign multiple variables in the same `mutate()` call, separated by commas (`,`)

```{r, echo = TRUE}
person_year_data %>%
  mutate(NextYear = Year + 1,
         Above100k = Income > 100000)
```

## case_when()

- A function that comes in handy a lot when using mutate to *create* a categorical variable is `case_when()`, which is sort of like `ifelse()` except it can cleanly handle way more than one condition
- Provide `case_when()` with a series of `if ~ then` conditions, separated by commas, and it will go through the `if`s one by one for each observation until it finds a fitting one. 
- As soon as it finds one, it stops looking, so you can assume anyone that satisfied an earlier condition doesn't count any more. Also, you can have the last `if` be `TRUE` to give a value for anyone who hasn't been caught yet

## case_when()

```{r, echo = TRUE}
person_year_data %>%
  mutate(IncomeBracket = case_when(
    Income <= 50000 ~ 'Under 50k',
    Income > 50000 & Income <= 100000 ~ '50-100k',
    Income > 100000 & Income < 120000 ~ '100-120k',
    TRUE ~ 'Above 120k'
  ))
```

## group_by()

- `group_by()` turns the dataset into a *grouped* data set, splitting each combination of the grouping variables
- Calculations like `mutate()` or (up next) `summarize()` or (if you want to get fancy) `group_map()` then process the data separately by each group

```{r, echo = TRUE}
person_year_data %>% group_by(Person) %>%
  mutate(Income_Relative_to_Mean = Income - mean(Income))
```

## group_by()

- It will maintain this grouping structure until you re-`group_by()` it, or `ungroup()` it, or `summarize()` it (which removes one of the grouping variables)
- How is this useful in preparing data? 
- Remember, we want to *look at where information is* and *think about how we can get it where we need it to be*
- `group_by()` helps us move information *from one row to another in a key variable* - otherwise a difficult move!
- It can also let us *change the observation level* with `summarize()`
- Tip: `n()` gives the number of rows in the group - handy! and `row_number()` gives the row number within its group of that observation

## summarize()

- `summarize()` *changes the observation level* to a broader level
- It returns only *one row per group* (or one row total if the data is ungrouped)
- So now your keys are whatever you gave to `group_by()`

```{r, echo = TRUE}
person_year_data %>%
  group_by(Person) %>%
  summarize(Mean_Income = mean(Income),
            Years_Tracked = n())
```

## summarize()

- `group_by() %>% summarize()` is the solution to many, many, many of your problems in data viz. If you're having a hard time doing something, ask yourself if `group_by() %>% summarize()` will do it for you
- To work right, `summarize()` needs to be given calculations to perform that return a single value. `summarize(mean_x = x)` won't take the mean of `x`, it'll just return the original data
- Protip in general: the mean of a binary variable is the proportion!

## rename()

- Lastly, and really only for aesthetic purposes, we have `rename()` for renaming your variables
- Often nice for making names manageable, or making them show up nicely in visualization (or especially table) output
- If your variable names have spaces or weird characters, be sure to put backticks around them

```{r, echo = TRUE}
person_year_data %>%
  rename(Name = Person, `Annual Income` = Income)
```

## Try It!

Using the same `construction` data from before (may have to reload it if you overwrote it before)

1. `mutate()` a new variable `avg` which adds Northeast, Midwest, South, and West, and then divides by 4
1. `rename()` that variable to Price Average (note the space in the name - backticks!)
1. `mutate()` a variable `firsthalf` that's "First half" for the first six months of the year and "Second half" for the second half (hint: `case_when()` and `row_number()`)
1. `group_by()` the `firsthalf` variable, and then `summarize()` to get the mean of `Price Average` by groups

## Try It!

```{r, echo = TRUE}
construction %>%
  mutate(avg = (Northeast + Midwest + South + West)/4) %>%
  rename(`Price Average` = avg) %>%
  mutate(firsthalf = case_when(row_number() <= 6 ~ "First Half", 
                               row_number() > 6 ~ "Second Half")) %>%
  group_by(firsthalf) %>%
  summarize(`Price Average` = mean(`Price Average`))
```

## group_by stuff

As I mentioned, students tend to forget that `group_by()` followed by `summarize()` or `mutate()` solves *so many problems*. Like what?

- "Find the (average / maximum / sum / first / etc.) for each (group)"
- Count the number of observations by group (`group_by(X) %>% summarize(n = n())`)
- Calculate the change each period by group (`arrange(date) %>% group_by(X) %>%` `mutate(change = Y - shift(Y))`)
- Really any time you want to do something repeatedly, once per group

# Factors, Dates, and Strings

## Manipulating Variables

- Those are the base **dplyr** verbs we need to think about
- They can be combined to do all sorts of things!
- But important in using them is thinking about what kinds of variable manipulations we're doing
- That will feed into our `mutate()`s and our `summarizes()`
- A lot of data cleaning is making an already-tidy variable usable!

## Variable Types

Three common variable types that need to be manipulated regularly in application to our data visualization tasks:

- Factors (categorical variables)
- Dates
- Strings

## Variable Types

- You can check the types of your variables by printing a `tibble()`, or `is.` and then the type, or doing str(data)
- You can generally convert between types using `as.` and then the type 

```{r, echo = TRUE}
taxdata %>%
  pivot_wider(names_from = 'TaxFormRow',
              values_from = 'Value') %>%
  mutate(Person = as.factor(Person),
         Income = as.numeric(Income),
         Deductible = as.numeric(Deductible),
         AGI = as.numeric(AGI))
```

## Factors

- The `factor()` function lets you specify these `labels`, and also specify the `levels` they go in - factors can be ordered! 
- When we get to graphing, setting the `levels` will determine the order in which things are graphed

```{r, echo = TRUE}
tibble(Income = c('50k-100k','Less than 50k', '50k-100k', '100k+', '100k+')) %>%
  mutate(Income = factor(Income, levels = c('Less than 50k','50k-100k','100k+'))) %>%
  arrange(Income)
```

## Factors

- (make sure you spell those level labels correctly, by the way!)
- Other neat tools for ordering factors:
- `reorder(x, Y)` will reorder the factor `x` according to the values of `Y`. Very handy for, say, ordering a set of bars on a bar graph by the height of the bars
- The entire **forcats** library, part of the **tidyverse**, has endless convenience functions for factors that I never bothered to learn much about


## Dates

- Dates are the scourge of data cleaners everywhere. They're just plain hard to work with!
- There are Date variables, Datetime variables, both of multiple different formats... eugh!
- **I strongly recommend** using the **lubridate** package whenever working with dates. See the [cheatsheet](https://github.com/rstudio/cheatsheets/raw/master/lubridate.pdf)
- If you're dealing with dates, before you work hard, stop and ask "oh I should use **lubridate**, that will make it easy"

## Convenient lubridate functions

- Convert strings to dates by telling it the order the information comes in: `ymd('2020-01-01')`, `ym('202001')`, `mdy('01-01-2020')`. 
- Combine that with `paste0()` (we'll get there with strings) to make a date out of multiple columns: `mutate(date = mdy(paste0(MONTH,DAY,YEAR)))`
- Get information about the date back out with `year(date)`, `quarter(date)`, `month(date)`, `day(date)`, etc.
- Label months with built-in R month names: `month.abb[month(date)]` or `month.name[month(date)]`

## Convenient lubriate functions

- Move around in time by adding or subtracting days/months/years: `ymd('2020-01-01') + years(1)`, `ymd('2020-01-01') + months(3)`, `ymd('2020-01-01') - days(7)`
- "Round down" the dates using `floor_date()`. `floor_date(ymd('2021-05-15'), 'month')` gives the first of May, 2021
- Note you can also work with `week()` of the year, or with date-time objects

## Characters/string

- Specified with `''` or `""`
- Use `paste0()` to stick stuff together! `paste0('h','ello', sep = '_')` is ''h_ello'`
- Note that `'\n'` can be used for line breaks
- Pasting together annotations and graph titles is a great idea!
- (btw you really don't want your numbers to be strings!! Don't put quotes around numbers!)

## stringr

- Just like **forcats** for categories and **lubridate** for dates, **stringr** (in the **tidyverse** as well) makes working with strings as easy as possible. See the [cheatsheet](https://raw.githubusercontent.com/rstudio/cheatsheets/main/strings.pdf)
- When working with textual data, first ask: hmm why don't I try looking in **stringr** to see if there's a relevant function?
- Conveniently most relevant functions start with `str_` so you can just type `stringr::str_` and let RStudio's autocomplete give some suggestions.

## Common String Manipulation Needs

The next few slides we're going to skip past, but they're here to come back to in case you need:

- A substring. `str_sub('hello',2,4)` is `'ell'`, `word('hello hi',2)` is `'hi'`
- Splitting strings in two with `str_split()` and `separate()`
- Decluttering strings and removing whitespace with `str_trim()/str_squish()`
- Fixing strings: `str_replace('MSFT stock','MSFT','Microsoft')` is `'Microsoft Stock'`
- Prettifying strings by making them title case, upper case, etc.


## Getting Substrings

- When working with things like nested IDs (for example, NAICS codes are six digits, but the first two and first four digits have their own meaning), you will commonly want to pick just a certain range of characters
- `str_sub(string, start, end)` will do this. `str_sub('hello', 2, 4)` is `'ell'`
- Note negative values read from end of string. `str_sub('hello', -1)` is `'o'`

## Getting Substrings

- Another way to pick out parts of a string is to get `word()`s from them
- `word('Sales Department', 1)` will give back just `'Sales'`

## Strings

- **Lots** of data will try to stick multiple pieces of information in a single cell, so you need to split it out! 
- Generically, `str_split()` will do this. `str_split('a,b', ',')[[1]]` is `c('a','b')`
- Often in already-tidy data, you want `separate()` from **tidyr**. Make sure you list enough new `into` columns to get everything!

```{r, echo = TRUE}
tibble(category = c('Sales,Marketing','H&R,Marketing')) %>%
  separate(category, into = c('Category1', 'Category2'), ',')
```

## Cleaning Strings

- Strings sometimes come with unwelcome extras! Garbage or extra whitespace at the beginning or end, or badly-used characters
- `str_trim()` removes beginning/end whitespace, `str_squish()` removes additional whitespace from the middle too. `str_trim(' hi  hello ')` is `'hi  hello'`.

## Cleaning Strings

- `str_replace` and `str_replace_all()` are often handy for eliminating (or fixing) unwanted characters

```{r, echo = TRUE}
str_replace('My name is Biarn', 'Biarn', 'Brian')
tibble(number = c('1,000', '2,003,124')) %>%
  mutate(number = number %>% str_replace_all(',', '') %>% as.numeric())
```

## Pretty Strings

- Stringr has a few functions for changing the case of strings, ideally to make them look nicer
- `str_to_upper()` and `str_to_lower()` can make all-caps or all-lowercase
- `str_to_title()` is very nice for making something title-case, Like This
- `str_to_sentence()` can easily convert something into sentence case, Like this.

## Detecting Patterns in Strings

- Often we want to do something a bit more complex. Unfortunately, this requires we dip our toes into the bottomless well that is *regular expressions*
- Regular expressions are ways of describing patterns in strings so that the computer can recognize them. Technically this is what we did with `str_replace_all(',','')` - `','` is a regular expression saying "look for a comma"
- There are a *lot* of options here. See the [guide](https://stringr.tidyverse.org/articles/regular-expressions.html)
- Common: `[0-9]` to look for a digit, `[a-zA-Z]` for letters, `*` to repeat until you see the next thing... hard to condense here. Read the guide.

## Detecting Patterns in Strings

- For example, some companies are publicly listed and we want to indicate that but not keep the ticker. `separate()` won't do it here, not easily!
- On the next page we'll use the regular expression `'\\([A-Z].*\\)'`
- `'\\([A-Z].*\\)'` says "look for a (" (note the `\\` to treat the usually-special ( character as an actual character), then "Look for a capital letter `[A-Z]`", then "keep looking for capital letters `.*`", then "look for a )"

## Detecting Patterns in Strings

```{r, echo = TRUE}
tibble(name = c('Amazon (AMZN) Holdings','Cargill Corp. (cool place!)')) %>%
  mutate(publicly_listed = str_detect(name, '\\([A-Z].*\\)'),
         name = str_replace_all(name, '\\([A-Z].*\\)', ''))
```

## For Example

- Let's remove the "county" from a list of county names 

```{r, echo = TRUE}
data("fips_to_names", package = 'SafeGraphR')

# Always check if the data is what you think! Do they all end in "County"?
# word(n) pulls the nth "word", -1 starts from the end
fips_to_names %>% pull(countyname) %>% word(-1) %>% table()
```

## For Example

- The ones we'd want to remove are probably "County" and "Parish"
- Let's use the easiest kind of regular expression: `|` to mean "or":

```{r}
fips_to_names <- fips_to_names %>%
  mutate(countyname = str_replace_all(countyname, ' County| Parish',''))
```

## Try It!

Load the `gss_cat` data. Then :

- Use `ym()` to create a `birthdate` variable equal to June 1 of the year in the `year` column, minus the number of years in the `age` column (`-years()`)
- Use `floor_date()` to set it back from June 1 of that year to January 1
- Set the levels of the `race` factor to Black, White, Other, in that order
- Use `table(gss_cat$rincome)` to look at the values income takes

(page 1!)

## Try It!

Then, if you're going into the string stuff, create `income_num` equal to the first number: 

1. use `word()` to get the first word, 
1. `str_sub()` or `str_replace_all()` to get rid of the "$", and
1. `as.numeric()` to make the result a number 

(you'll get some NAs for the non-numbers, and let's ignore the "Lt $1000" case we miss for now)

## Try It!

```{r, echo = TRUE, eval = FALSE}
data(gss_cat)
gss_cat %>%
  mutate(birthdate = ym(paste0(year, '06')) - years(age)) %>%
  mutate(birthdate = floor_date(birthdate, 'year')) %>%
  mutate(race = factor(race, levels = c('Black','White','Other'))) %>%
  mutate(income_num = rincome %>% word(1) %>% str_sub(2) %>% as.numeric())
```

```{r, echo = FALSE, eval = TRUE}
data(gss_cat)
gss_cat %>%
  mutate(birthdate = ym(paste0(year, '06')) - years(age)) %>%
  mutate(birthdate = floor_date(birthdate, 'year')) %>%
  mutate(race = factor(race, levels = c('Black','White','Other'))) %>%
  mutate(income_num = rincome %>% word(1) %>% str_sub(2) %>% as.numeric()) %>%
  datatable()
```

# Automation

## Automation

- Data cleaning is often very repetitive
- You shouldn't let it be!
- Not just to save yourself work and tedium, but also because standardizing your process so you only have to write the code *once* both reduces errors and means that if you have to change something you only have to change it once

## Breezing Past

We've already gone through a lot! Just remember: **if you're doing the same thing a bunch of times in a row, please don't copy/paste or actually type it all out. Come back to these slides**. That includes:

- Doing the same `summarize/mutate` calculation to a lot of different variables (use `across()`)
- Reusing the same few lines of code (or big chunk of code) over and over in different parts of your analysis (writing functions)
- Performing the same actions to many different objects or elements of a vector (**purrr**)

## across()

- If you have a lot of variables, cleaning them all can be a pain. Who wants to write out the same thing a million times, say to convert all those read-in-as-text variables to numeric?
- Old versions of **dplyr** used "scoped" variants like `mutate_at()` or `mutate_if()`. As of **dplyr 1.0.0**, these have been deprecated in favor of `across()`
- `across()` lets you use all the variable-selection tricks available in `select()`, like `starts_with()` or `a:z` or `1:5`, but then lets you apply functions to each of them in `mutate()` or `summarize()`
- similarly `rowwise()` and `c_across()` lets you do stuff like "add up a bunch of columns"

## across()

- `starts_with('price_growth')` is the same here as `4:5` or `c(price_growth_since_march_4, price_growth_daily)`

```{r, echo = TRUE}
stockgrowth <- tibble(ticker = c('AMZN','AMZN', 'AMZN', 'WMT', 'WMT','WMT'),
       date = as.Date(rep(c('2020-03-04','2020-03-05','2020-03-06'), 2)),
       stock_price = c(103,103.4,107,85.2, 86.3, 85.6)) %>%
  arrange(ticker, date) %>% group_by(ticker) %>%
  mutate(price_growth_since_march_4 = stock_price/first(stock_price) - 1,
         price_growth_daily = stock_price/lag(stock_price, 1) - 1) 
stockgrowth %>%
  mutate(across(starts_with('price_growth'), function(x) x*10000)) # Convert to basis points
```

## Writing Functions

- Generally, **if you're going to do the same thing more than once, you're probably better off writing a function**
- Reduces errors, saves time, makes code reusable later!

```{r, echo = TRUE, eval = FALSE}
function_name <- function(argument1 = default1, argument2 = default2, etc.) {
  some code
  result <- more code
  return(result)
  # (or just do result by itself - the last object printed will be automatically returned if there's no return())
}
```

## Writing Functions

The same basis-point conversion from before:

```{r, echo = TRUE}
do_bps <- function(p) {
  bps <- p*10000
  return(bps)
}
stockgrowth %>%
  mutate(across(starts_with('price_growth'), do_bps)) # Convert to basis points
```


## purrr

- One good way to apply functions iteratively (yours or not) is with the `map()` functions in **purrr**
- `map()` usually generates a `list()`, `map_dbl()` a numeric vector, `map_chr()` a character vector, `map_df()` a `tibble()`...
- It iterates through a `list`, `data.frame/tibble` (which are technically `list`s, or `vector`, and then applies a function to each of the elements

```{r, echo = TRUE}
person_year_data %>%
  map_chr(class)
```

## purrr

- Or looping more generally for diagnostic or wrangling purposes. Perhaps you have a `summary_profile()` function you've made, and want to check each state's data to see if its data looks right. You could do 

```{r, echo = TRUE, eval = FALSE}
data %>% pull(state) %>% unique() %>% map(summary_profile)
```

## purrr

- You can use it generally in place of a `for()` loop

```{r, echo = TRUE, eval = FALSE}
my_function <- function(i) {
  print(i+1)
}
# Print the numbers 1:10
1:10 %>% map(my_function)
```

- See the [purrr cheatsheet](https://github.com/rstudio/cheatsheets/blob/main/purrr.pdf)

# Finishing Up, and an Example!

## Some Final Notes

- Importing and exporting data is way easier than it used to be, with **rio**
- `import()` for any filename will import it (watch your working directory!)
- `import_list()`, given a list of filenames or a single Excel workbook with multiple sheets, will import them all to a list (access each one with `[[1]]`, `[[2]]`, etc., or if they're row-bindable, just stack them all on top of each other into one data set with `rbind = TRUE`)
- Then `export()` to save when you're done
- Note: `import()` can get a file that's currently inside a `.zip` to save space, or `export()` to `.zip`

## Data File Types

- `.Rdata`: very compact, and flexible (can contain more than just one data set), but specific to R and can't be used most other places
- `.csv`: the most general and can be used by anything, but large filesizes and you lose some data detail (like, factors don't carry over)
- `.parquet`: very compact and also fairly general, but also new and not everyone will know what to do with it

## Some Final Notes

- Also, please, please, *please* **DOCUMENT YOUR DATA**
- At the very least, keep a spreadsheet/\code{tibble} with a set of descriptions for each of your variables
- Also look into the **sjlabelled** or **haven** packages to add variable labels directly to the data set itself
- Once you have your variables labelled, `vtable()` in **vtable** can generate a documentation file for sharing

## Let's Try It!

1. Write a function that takes `x` and `a` and returns `x+a`
1. Load the `mtcars` data.
1. Use `mutate(across())` to add 1 to all columns ending in 'p', using the function you wrote
1. Put steps 2 and 3 inside another function that takes `a` as an argument, change "add 1" to "add `a`" and returns the resulting data
1. Use `map_df` to stack together data sets with `a` values from 1 to 5
1. Save the result as a CSV with `export()` from **rio**

## Let's Try It!

```{r, eval = FALSE, echo = TRUE}
library(rio)
x_plus_a <- function(x, a) { x + a }

add_to_mtcars <- function(a) {
  data(mtcars)
  adjusted <- mtcars %>% mutate(across(ends_with('p'), \(x) x_plus_a(x,a)))
  return(adjusted)
}

1:5 %>% map_df(add_to_mtcars) %>% export('myfile.csv')
```

```{r, echo = FALSE, eval = TRUE}
x_plus_a <- function(x, a) { x + a }

add_to_mtcars <- function(a) {
  data(mtcars)
  adjusted <- mtcars %>% mutate(across(ends_with('p'), x_plus_a, a = a))
  return(adjusted)
}

1:5 %>% map_df(add_to_mtcars) %>% datatable()
```

